The elevational structure of a mountain mutualism: richness, abundance, and β-diversity of bumble bees and their floral hosts

### 2021-09-24 code/ms1_beta_diversity.Rmd
almost there...

# For each site pair, one will be higher and the other lower; thus, when their order is reversed only the sign of the difference changes. Since no two sites have exactly the same elevation, `filter(elev.diff >= 0)` deduplicates the data by dropping the ordering of a given pair of sites (one of the two orderings) that yields a negative elevation difference while retaining the alternative ordering that yields a positive elevation difference. By >= 0, I retain comparisons of a site to itself. I drop (in the ggplot call) within-site comparisons when I plot against elevation difference, and I isolate within-site comparison when I plot against time. In the latter case, it is also necessary to deduplicate, since each pair of site dates occurs in A-B and B-A. For that, I filter (again, in the ggplot call) so that yday.diff > 0, similarly dropping the orderings that yield negative differences. Note that all this deduplication nonsense is not necessary for the betalinkr call because it deduplicates internally.

# Fun with betapart
fl_betapart <- survey %>%
  group_by(site, plant.sp) %>%
  summarize(count = sum(flower.cover)) %>%
  mutate(present = ifelse(count > 0, 1, 0)) %>%
  select(-count) %>%
  pivot_wider(names_from = plant.sp, values_from = present) %>%
  replace(is.na(.), 0) %>%
  column_to_rownames("site") %>%
  beta.pair(index.family = "jaccard")

fl_betapart.turnover <- fl_betapart$beta.jtu %>%
  as.matrix() %>%
  as.data.frame() %>%
  rownames_to_column(var = "site1") %>%
  pivot_longer(cols = -site1, names_to = "site2", values_to = "tu") %>%
  filter(site1 != site2) %>%
  left_join(site_data, by = c("site1" = "site")) %>%
  dplyr::select(site1, site2, tu, tree_line1 = tree_line) %>%
  left_join(site_data, by = c("site2" = "site")) %>%
  dplyr::select(site1, site2, tu, tree_line1, tree_line2 = tree_line)

fl_betapart.nestedness <- fl_betapart$beta.jne %>%
  as.matrix() %>%
  as.data.frame() %>%
  rownames_to_column(var = "site1") %>%
  pivot_longer(cols = -site1, names_to = "site2", values_to = "ne") %>%
  filter(site1 != site2) %>%
  left_join(site_data, by = c("site1" = "site")) %>%
  dplyr::select(site1, site2, ne, tree_line1 = tree_line) %>%
  left_join(site_data, by = c("site2" = "site")) %>%
  dplyr::select(site1, site2, ne, tree_line1, tree_line2 = tree_line)

fl_betapart.collate <- full_join(fl_betapart.turnover, fl_betapart.nestedness) %>%
  mutate(tree_line.comp = case_when(
    tree_line1 == "below" & tree_line2 == "below" ~ "bb",
    tree_line1 == "above" & tree_line2 == "above" ~ "aa",
    tree_line1 != tree_line2 ~ "ab"
  )) %>%
  pivot_longer(cols = c(tu, ne), values_to = "value", names_to = "var")

ggplot(fl_betapart.collate, aes(var, value)) +
  geom_boxplot() +
  facet_wrap(~tree_line.comp)
```

>>>>

# For each site pair, one will be higher and the other lower; thus, when their order is reversed only the sign of the difference changes. Since no two sites have exactly the same elevation, `filter(elev.diff >= 0)` deduplicates the data by dropping the ordering of a given pair of sites (one of the two orderings) that yields a negative elevation difference while retaining the alternative ordering that yields a positive elevation difference. By >= 0, I retain comparisons of a site to itself. I drop (in the ggplot call) within-site comparisons when I plot against elevation difference, and I isolate within-site comparison when I plot against time. In the latter case, it is also necessary to deduplicate, since each pair of site dates occurs in A-B and B-A. For that, I filter (again, in the ggplot call) so that yday.diff > 0, similarly dropping the orderings that yield negative differences. Note that all this deduplication nonsense is not necessary for the betalinkr call because it deduplicates internally.
```

### 2021-09-24 ms_1/ms_revision1.Rmd
almost there...

---
title: "The elevational structure of a mountain mutualism: modes and thresholds in the responses of bumble bees, wildflowers, and their interactions"

@@

@@

Bumble bees (Hymenoptera: *Bombus* spp.) and their floral hosts provide an excellent system in which to study elevational patterns at both the species- and interaction-level. Bumble bees are believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary, and their subsequent spread and diversification along mountain corridors and into lowland habitats was likely driven by alternating range expansion during cooling periods and retreat to higher elevations during warming periods [@Hines2008-er; @Stewart2010-jn; @Martinet2018-rb]. Today, they remain characteristically cold-adapted organisms [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes [@Williams1998-xi]. In these habitats, they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Goulson2010-lo]. As in the past, so now are mountain habitats critical theaters of ecological change for bumble bees. Warming climates are expected to cause habitat loss and potentially novel competitive pressures [@Marshall2020-pc; @Soroye2020-zb; though see @Guzman2021-my]. At the same time, socioeconomic drivers are causing the reforestation of pastures on mountain slopes and the intensification of agriculture in lowlands and valleys, both of which entail a loss of floral resources for bumble bees [@Tasser2002-om; @Tappeiner2006-jp; @Tasser2007-qo; @Walcher2017-kw].

@@

The response of bumble bee richness to elevation exhibited an approximately linear increase up to 1600-1700 m and then approximately leveled off **(Figure 3A)**. Neither mowing nor abandonment had a significant effect on abundance relative to the reference management level of grazing (p > 0.05), but richness was significantly lower in 2012 than in the reference year of 2010 (p = 0.04) **(Table 1)**. In contrast, the response of floral richness exhibited a clear unimodal pattern, peaking between 1300-1500 m **(Figure 3B)**. Management had no significant effect on floral richness (P > 0.1), but floral richness was significantly lower in 2011 (p < 0.001) and 2012 (p = 0.03) than in 2010. The response of interaction richness to elevation was intermediate to that of bumble bee and floral richness, peaking around 1500 m and leveling off at higher elevation **(Figure 3C)**. Management did not have a significant effect on interaction richness (p > 0.1), but interaction richness was significantly higher in 2011 (p < 0.05) and lower in 2012 (p < 0.001) relative to the reference level of 2010.

@@

At the species level, bumble bees exhibited variable and sometimes divergent responses to elevation **(Figure 5)**. *Bombus pyrenaeus*, *B. mendax*, *B. monticola*, and *B. mucidus* exhibited approximately linear increases with elevation (p < 0.05). In contrast, *Bombus psithyrus* and *B. pascuorum* exhibited approximately linear decreases with elevation (p < 0.001). *Bombus pratorum* was unique in exhibiting a strongly unimodal response to elevation peaking around 1600 m (p < 0.001). The remaining species did not respond significantly to elevation (p > 0.05).

@@

Bumble bees, flora, and their interactions exhibited marked differences in their rates of β-diversity in response to elevation. **(Figure 4)**. Bumble bee turnover was relatively high from ~600-1100 m, briefly leveled off, and then accelerated again around 1250 m. Floral Turnover was high from ~600-1000 m, gradually leveled off from ~1000-1500 m, and then accelerated again around the tree line ecotone at ~1500 m. Interaction turnover mirrored the high rates of bumble bee and floral β-diversity from ~600-1000 m, but it exhibited a markedly different pattern above 1000 m, with relatively low turnover between 1000-1750 m and then a sharp acceleration from ~1750-2000 m. The conditional effect of elevation was significant (p < 0.001) in all models, and its importance was 4-20 time higher that of sampling intensity and geographic distance **(Table 3)**. Nevertheless, geographic distance was a significant predictor in all models (p < 0.001), and sampling intensity was a significant predictor in the bumble bee (p = 0.004) and interaction models (p < 0.001). Our GDMs explained 31% of total deviance in bumble bee β-diversity, 73% of floral β-diversity, and 71% of interactions β-diversity.

@@

## Big picture

The core finding of our study, evident in each of our analyses, is that a linear elevation gradient can generate complex and nonlinear biological responses in a mountain bumble- bee-wildflower community in terms of richness, abundance and β-diversity, and at the level of both species and interactions.

The positive response of bumble bee richness to elevation sets bumble bees apart from most other taxa [@Rahbek1995-oj; @Guo2013-tu; @Peters2016-hf], underscoring their exceptional adaptedness to mountain environments. On higher mountains, the increase of bumble bee richness with elevation eventually gives way to a unimodal response, but peak richness is often found above 3000 m [e.g. @Williams1991-zu; @Streinzer2019-bz], ~1000 m above the highest peaks in the Berchtesgadener Alps. Thus, in our study system, the positive response of bumble bee richness to elevation encompasses the entire mountain slope, in agreement with the findings of @Minachilis2020-xd along the comparable elevation gradient (300---2700 m) of Mt. Olympus. Importantly, the positive response of bumble bee richness stands in sharp contrast to the monotonic decline of overall bee richness with elevation in the Berchtesgadener Alps [@Hoiss2012-sn], highlighting the special importance of bumble bees as pollinators where few other bees persist.

While the response of overall bumble bee abundance to elevation was approximately flat, this aggregate pattern concealed divergent species-level responses. The strongly positive responses of *B. mendax*, *B. monticola*, *B. mucidus*, *B. pyrenaeus* are consistent with the known affinity of these species for high-elevation and/or high latitude habitats [@Rasmont2010-ci]. At the other extreme, *B. pascuorum* and *B. psithyrus* exhibited a strongly negative responses to elevation. In the case of *B. pascuorum*, this pattern is likely driven by its exceptionally long nesting cycle, which is a liability under the short growing seasons associated with higher elevations [@Neumayer2009-gx]. Growing season length may also limit the abundance of *B. psithyrus* at high elevation, since cuckoo bumble bees emerge later than the queens of their host species and must successfully usurp host nests before they can begin to raise brood [@Goulson2010-lo]. *Bombus pratorum* was unique in exhibiting a strongly unimodal abundance pattern with a well-defined peak between 1500-1600 m, . @Iserbyt2008-jn describe a similar abundance pattern of *B. pratorum* in the Eyne Valley of the Pyrenees.

The roughly sigmoidal pattern of β-diversity rate in bumble bees across our elevation gradient, revealed by our GDM analysis, reflects the reciprocal abundance patterns of high- and low-elevation species described above as well as the occurrence of certain rarer species only at middle-elevation sites. It is important to note, though, that bumble bee turnover was low over all, and 8 out of the 16 species spanned the entire elevation gradient (Appendix 1).

Floral β-diversity also exhibited a sigmoidal pattern, but its inflection point clearly corresponds to the tree line ecotone at ~1500 m. This floristic importance of the tree line has been noted in previous work [e.g. @Pellissier2010-kg; @Descombes2017-id], and our study corroborates these findings specifically for the subset of flora visited by bumble bees. Interestingly, the tree line does not constitute a discrete transition to a new equilibrium, but rather a threshold at which turnover begins a process of acceleration that holds for the remainder of the elevation gradient. While the most obvious effect of the tree line is the relatively discrete transition from forest to scrub/grassland, the tree line also initiates a process of angiosperm species attrition that proceeds as a rate of about 40 species per 100 m [@Korner2003-sc], and the pattern of increasing turnover above the tree line mirrors the patterns of decreasing species richness **(Figure 3)** and floral abundance **(Figure 4)**. Thus, the bumble bees foraging above the tree line must cope with a floral community that not only turns over rapidly in composition but also declines steeply in diversity and abundance.

We saw a noteworthy contrast between the relative stability of the bumble bee community and the extreme dynamism of the floral community. The dominance of total interaction β-diversity by floral species turnover in our system is consistent with the findings of @Simanonok2014-ou in the Rocky Mountains of North America, and it is perhaps unsurprising given the exceptional cold-hardiness of bumble bees and the sensitivity of plants to temperature gradients. Nevertheless, it highlights the behavioral challenge that bumble bees have presumably faced for the entirety of their evolutionary history as mountain florivores: how to forage efficiently in a diverse floral community that exhibits rapid spatial and temporal turnover. A fascinating question that could not be answered by our sampling approach is whether mountain bumble bees adaptively forage up- or downslope, as suggested by @Lundberg1980-qk. Bumble bees have large foraging ranges and have been shown both to track resources through space and time [@Devoto2014-ay] and to cross forest matrix to reach patches of foraging habitat [@Mola2020-yc]. Assuming an average slope of 20° and a foraging range of 1 km, a bumble bee could travel up- or downslope by more than 340 m, thus spanning an elevation band nearly 700 m wide. Such 3-dimensional foraging would enable bumble bees both to exploit the elevational turnover of floral species and to track preferred species through their elevationally staggered phenology, the latter constituting a sort of physiological time-travel [@Van_Straalen1983-df].

<!-- Above the tree line, we found that flora resource availability declines sharply as flower-rich grassland and shrubland transitions to sparse alpine scree. The observation that the upslope advancement of alpine tree lines often lags behind climate warming [@Dullinger2004-qu] suggests that high elevation bumble bees could move upslope faster than their floral resources and become increasingly confined to resource-poor habitat, though the opposite pattern could also occur if the climate-tracking of bumble bees is limited by other factors, such as nest site availability [@Marshall2020-pc].  -->

<!-- Several studies have found that the introduction or removal of bumble bee species from a community can have strong effects on the fitness of other species and on patterns of floral visitation [@Nagamitsu2010-wi; @Brosi2013-mu; @Ishii2013-sm]. If elevational range shifts introduce new competitive pressures via the encroachment of low-elevation species into the historic ranges of high-elevation species . . . -->

<!-- The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective [@Soroye2020-zb], but they also provide a model system in which to study the basic process of community assembly under climate change [@Hoiss2012-sn; @Classen2015-jc]. -->

# ACKNOWLEDGEMENTS

@@

**Table 2:** Tabular output of species-level abundance GAMs. Parametric model terms are summarized with an estimated intercept effect ("estimate") while smooth terms are summarized with their effective degrees of freedom ("edf"), the latter representing the complexity (i.e. "wiggliness") of the fitted spline.

@@

**Table 3:** Summary of GDM results for each model (bumble bees, flora, and interactions, respectively). Variable importance represents the percent of explained deviance lost when the focal variable is permuted, and it is mapped to a color ramp to aid interpretation. The p-value of each variable (Var.pval) is based on comparing the deviance explained by the unpermuted model (Model.dev.exp) to the distribution of deviance explained when the focal variable is permuted. Similarly, the model p-value (Model.pval) is based on a comparison between the deviance explained by the unpermuted model and the distribution of deviance explained when the entire predictor matrix is permuted.

>>>>

---
title: "The elevational structure of a mountain mutualism: richness, abundance, and β-diversity of bumble bees and their floral hosts"

@@

Environmental gradients generate and maintain biodiversity on Earth. Mountain slopes are among the most pronounced terrestrial environmental gradients, and the elevational structure of species and their interactions can provide unique insight into the processes that govern community assembly and function in mountain ecosystems.

We recorded bumble- bee-flower interactions over three years along an 1400 m elevational gradient in the German Alps. Using nonlinear modeling, we analyze the elevational patterns at the nested levels of species abundance, species β-diversity, and interaction β-diversity.

While floral richness exhibited mid-elevation peak, bumble bee richness increased with elevation before leveling off at the highest sites, demonstrating the exceptional adaptation of these bees to cold temperatures and short growing seasons. At a species level, though, bumble bees exhibited divergent responses to elevation in terms of abundance, with a clear separation between species preferring low vs. high elevations.

@@

Bumble bees (Hymenoptera: *Bombus* spp.) and their floral hosts provide an excellent system in which to study elevational patterns at both the species- and interaction-level. Bumble bees are believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary, and their subsequent spread and diversification along mountain corridors and into lowland habitats was likely driven by alternating range expansion during cooling periods and retreat to higher elevations during warming periods [@Hines2008-er; @Stewart2010-jn; @Martinet2018-rb]. Today, they remain characteristically cold-adapted organisms [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes [@Williams1998-xi]. In these habitats, they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Goulson2010-lo]. As in the past, so now are mountain habitats critical theaters of ecological change for bumble bees. Warming climates are expected to cause habitat loss and potentially novel competitive pressures [@Kerr2015-dg; @Soroye2020-zb; though see @Guzman2021-my]. At the same time, socioeconomic drivers are causing the reforestation of pastures on mountain slopes and the intensification of agriculture in lowlands and valleys, both of which entail a loss of floral resources for bumble bees [@Tasser2002-om; @Tappeiner2006-jp; @Tasser2007-qo; @Walcher2017-kw].

@@

The response of bumble bee richness to elevation exhibited an approximately linear increase up to 1600-1700 m and then leveled off and declined slightly at the upper limit of the gradient **(Figure 3A)**. Neither mowing nor abandonment had a significant effect on abundance relative to the reference management level of grazing (p > 0.05), but richness was significantly lower in 2012 than in the reference year of 2010 (p = 0.04) **(Table 1)**. In contrast, the response of floral richness exhibited a clear unimodal pattern, peaking between 1300-1500 m **(Figure 3B)**. Management had no significant effect on floral richness (P > 0.1), but floral richness was significantly lower in 2011 (p < 0.001) and 2012 (p = 0.03) than in 2010. The response of interaction richness to elevation was intermediate to that of bumble bee and floral richness, peaking around 1500 m and leveling off at higher elevation **(Figure 3C)**. Management did not have a significant effect on interaction richness (p > 0.1), but interaction richness was significantly higher in 2011 (p < 0.05) and lower in 2012 (p < 0.001) relative to the reference level of 2010.

@@

At the species level, bumble bees exhibited variable and sometimes divergent responses to elevation **(Figure 5; Table 3)**. *Bombus pyrenaeus*, *B. mendax*, *B. monticola*, and *B. mucidus* exhibited approximately linear increases with elevation. In contrast, *Bombus psithyrus* and *B. pascuorum* exhibited approximately linear decreases with elevation. *Bombus pratorum* was unique in exhibiting a strongly unimodal abundance pattern with a well-defined peak between 1500-1600 m, approximately at the tree line ecotone, similar to the abundance pattern of *B. pratorum* described by @Iserbyt2008-jn in the Eyne Valley of the Pyrenees. The remaining species did not respond significantly to elevation (p > 0.05).

@@

Bumble bees, flora, and their interactions exhibited marked differences in their rates of β-diversity in response to elevation. **(Figure 4)**. Bumble bee turnover was relatively high from ~600-1100 m, briefly leveled off, and then accelerated again around 1250 m. Floral Turnover was high from ~600-1000 m, gradually leveled off from ~1000-1500 m, and then accelerated again around the tree line ecotone at ~1500 m. Interaction turnover mirrored the high rates of bumble bee and floral β-diversity from ~600-1000 m, but it exhibited a markedly different pattern above 1000 m, with relatively low turnover between 1000-1750 m and then a sharp acceleration from ~1750-2000 m. The conditional effect of elevation was significant (p < 0.001) in all models, and its importance was 4-20 time higher that of sampling intensity and geographic distance **(Table 4)**. Nevertheless, geographic distance was a significant predictor in all models (p < 0.001), and sampling intensity was a significant predictor in the bumble bee (p = 0.004) and interaction models (p < 0.001). Our GDMs explained 31% of total deviance in bumble bee β-diversity, 73% of floral β-diversity, and 71% of interactions β-diversity.

@@

Our study revealed strong and sometimes contrasting patterns of elevational structure in the richness, abundance, and β-diversity of mountain bumble bees and wildflowers. By extending our analysis to the level of species interactions, we show how the elevational patterns within bumble bee and wildflower communities are propagated to the functional relationships between these two guilds. These patterns suggest underlying processes governing the assembly of the bumble- bee-wildflower networks in mountain habitats, shedding light on how this mutualism has evolved in the past and how it will respond to the nascent forces of climate and land use change.

While floral richness exhibited the expected unimodal response to elevation, bumble bee richness responded positive to elevation for nearly the whole gradient, showing only a slight declining trend at the highest sites. The positive response of bumble bee species richness to elevation sets bumble bees apart from most other taxa [@Rahbek1995-oj; @Guo2013-tu; @Peters2016-hf], underscoring their exceptional adaptation to mountain environments. On higher mountains, the increase of bumble bee richness with elevation eventually gives way to a unimodal response, but peak richness is often found above 3000 m [e.g. @Williams1991-zu; @Streinzer2019-bz], 1000 m above the highest peaks in the Berchtesgadener Alps. Thus, in our study system, the positive response of bumble bee richness to elevation encompasses almost the entire mountain slope, in agreement with the findings of @Minachilis2020-xd along a comparable elevation gradient (300-2700 m) at Mt. Olympus, Greece. The positive response of bumble bee richness stands in sharp contrast to the monotonic decline of overall bee richness with elevation in the Berchtesgadener Alps [@Hoiss2012-sn], highlighting the special importance of bumble bees as pollinators where few other bees persist.

The response of overall bumble bee abundance to elevation was approximately flat, but this aggregate pattern concealed divergent species-level responses. The strongly positive responses of *B. mendax*, *B. monticola*, *B. mucidus*, *B. pyrenaeus* are consistent with the known affinity of these species for high-elevation habitats [@Rasmont2010-ci]. At the other extreme, *B. pascuorum* and *B. psithyrus* exhibited a strongly negative responses to elevation. In the case of *B. pascuorum*, this pattern is likely driven by its exceptionally long nesting cycle, which limits its fitness in the short growing seasons of higher elevations [@Neumayer2009-gx]. Growing season length may also explain the lower abundance of *B. psithyrus* at high elevation, since cuckoo bumble bees emerge later than the queens of their host species and must successfully usurp host nests before they can begin to raise brood [@Goulson2010-lo]. Nevertheless, while the divergent responses of bumble bee species to elevation are informative, it is important to emphasize the general robustness of bumble bees to elevational variation. Nine out of the 16 species in our study system did not respond significantly to elevation, and 8 out of 16 were observed across the full span of the gradient.

The relative stability of the bumble bee community in our study stands in stark contrast to the extreme dynamism of the floral community, as illustrated by the dominance of total interaction β-diversity by floral species turnover. This pattern is consistent with the findings of @Simanonok2014-ou in the Rocky Mountains of North America, and it is perhaps unsurprising given the exceptional cold-hardiness of bumble bees and the sensitivity of plants to temperature gradients. Nevertheless, it highlights the behavioral challenge that bumble bees have presumably faced for the entirety of their evolutionary history as mountain florivores: how to forage efficiently in a diverse floral community that exhibits rapid spatial and temporal turnover. A fascinating question that could not be answered by our sampling approach is whether mountain bumble bees actively exploit the elevational sensitivity of floral communities by adaptively foraging up- or downslope, as suggested by @Lundberg1980-qk. Bumble bees have large foraging ranges and have been shown both to track resources through space and time [@Devoto2014-ay] and to cross forest matrix to reach patches of foraging habitat [@Mola2020-yc]. Assuming an average slope of 20° and a foraging range of 1 km, a bumble bee could travel up- or downslope by more than 340 m, thus spanning an elevation band nearly 700 m wide. Such 3-dimensional foraging would enable bumble bees both to exploit the elevational turnover of floral species and to track preferred species through their elevationally staggered phenology, the latter constituting a sort of physiological time-travel [@Van_Straalen1983-df].

The roughly sigmoidal pattern of bumble bee β-diversity rate in response to elevation, revealed by our GDM analysis, probably reflects the reciprocal abundance patterns of high- and low-elevation species described above as well as the occurrence of certain rarer species only at middle-elevation sites. As discussed above, though, bumble bee turnover was low over all, demonstrating the robustness of bumble bees to the environmental variation encompassed by our elevation gradient. Floral β-diversity rate also exhibited a sigmoidal pattern, and its inflection point coincides with the tree line ecotone at ~1500 m. The floristic importance of the tree line has been noted in previous work [e.g. @Pellissier2010-kg; @Descombes2017-id], and our study corroborates these findings specifically for the subset of flora visited by bumble bees. Interestingly, the tree line did not constitute a discrete transition to a new equilibrium fo floral composition, but rather a threshold marking the beginning of a process of β-diversity acceleration that holds for the remainder of the elevation gradient. While the most obvious effect of the tree line may be the relatively discrete transition from forest to scrub/grassland, the tree line also initiates a process of angiosperm species attrition that proceeds at a rate of about 40 species per 100 m [@Korner2003-sc]. Indeed, the accelerated β-diversity we observed above the tree line mirrors simultaneous patterns of decreasing floral richness and abundance **(Figures 3B, 4B)**.

In contrast, the rate of interaction β-diversity exhibited a late and very pronounced acceleration at ~1800 m, indicating that the two highest sites (at 1984 m and 2032 m) differed strongly in interaction composition from those ~200 m below them. At these highest sites, the growing season is very short and vegetation is sparse, giving way to bare rock and scree landscapes, and it may be that the last 200 m interval of our elevation gradient imposes the most stringent environmental filter. It is interesting that interaction β-diversity did not exhibit the inflection point at the tree line ecotone (~1500 m) that we saw for floral β-diversity, despite the fact that overall interaction β-diversity was driven primarily by floral β-diversity. A possible explanation for this pattern is that there could be an elevational lag in bumble bee floral selection such that most bumble bee species do not switch to supra-tree-line floral species when these species first appear, but rather maintain associations with flora whose ranges span the tree line until the extreme species turnover between 1800 and 2000 m forces them to adopt new floral hosts.

The poverty of floral resources at the highest elevations raises the question of why bumble bees choose to forage at such elevations at all. For some species, the selection of high elevation habitats may be influenced by thermal tolerances [@Oyen2016-ng], but that fails to explain the presence of elevation-generalist species (including the majority of the species in our study system) in resource-poor high-elevation habitats. It is possible that nest site limitation at lower elevations drives some bumble bees to establish nests at high elevation, and their ability to forage downslope is subsequently constrained by their flight range, which may be reduced due to the energetic costs of flight in cold temperatures [@Kovac2015-wy]. It is also possible that high elevations, while sparse in absolute floral resources, offer an escape from more competitive conditions at lower elevations where forager density is higher [@Hoiss2015-gv; @Classen2020-nh]. Similarly, higher elevations could also offer an escape from density-dependent parasite and pathogen pressure [@Parsche2018-ar]. Finally, it is worth speculating that an affinity for open, high-elevation habitats might be an evolutionary legacy in mountain bumble bees, since the existence of flower-rich meadows in the forest matrix below the tree line is largely an artifact of human land use arising mainly in the last 500 years of bumble bees' evolutionary history [@Hejcman2013-ca].

Mountains are unique ecological theaters in which the extrinsic forces of climate and land use change interact with intrinsic environmental gradation to shape the ecology of species and their interactions [@Peters2019-km]. Given the elevational patterns of richness, abundance, and β-diversity among bumble bees and wildflowers in our study, what can be said about the response of these communities, and their functional relationships, to climate and land use change? One of the more troubling observations in our study is that the bulk of floral resources were located at low- to mid-elevation. If the process of mid-elevation pasture abandonment continues, the flower-rich meadows that now perforate the forest matrix will eventually close into flower-poor coniferous forest, and bumble bees will be pushed either into the more agriculturally intensified land at lower elevation or the relatively resource-poor open habitat above the tree line. This land- use-driven redistribution of bumble bees will interact with the concomitant process of climate warming, which has already generated measurable upslope shifts in some bumble bees species [@Ploquin2013-da; @Pyke2016-ll; @Marshall2020-pc], sometimes resulting in increased species richness at higher elevations [@Fourcade2019-ct]. Under this scenario, the high-elevation bumble bees in the Berchtesgadener Alps can be expected to face increased densities of low-elevation or elevation-generalist congeners, and the introduction of competing bumble bee species in a given habitat can have strong effects both on the fitness of their congeners and on patterns of floral visitation [@Nagamitsu2010-wi; @Brosi2013-mu; @Ishii2013-sm]. Nevertheless, the tolerance of bumble bees for a relatively broad climate envelope together with their long flight range may enable them to cope with climate and land use change better than their floral mutualists. The flowering plant community in our study system was strongly structured by elevation, and there is evidence that the upslope movement of plants in response to warming temperatures may outpace that of bumble bees [@Marshall2020-pc; though see @Pyke2016-ll]. Moreover, it is the plant community that experiences the direct effects of pasture abandonment at mid-elevation and agricultural intensification at low-elevation.

Finally, our study has focused on contemporary elevational patterns of mountain bumble bees and wildflowers, explored the processes that might regulate them, and related these processes to the nascent pressures of climate and land use change. Mountains are, however, inherently dynamic systems, and have been such throughout the co-evolutionary history of bumble bees and their floral mutualists. Both contemporary patterns and potential responses to global change processes would be elucidated by greater integration with the historical biogeography of bumble bees and wildflowers and their responses to previous episodes of global change [e.g. @Martinet2018-rb].

# ACKNOWLEDGMENTS

@@

**Table 3:** Tabular output of species-level abundance GAMs. Parametric model terms are summarized with an estimated intercept effect ("estimate") while smooth terms are summarized with their effective degrees of freedom ("edf"), the latter representing the complexity (i.e. "wiggliness") of the fitted spline.

@@

**Table 4:** Summary of GDM results for each model (bumble bees, flora, and interactions, respectively). Variable importance represents the percent of explained deviance lost when the focal variable is permuted, and it is mapped to a color ramp to aid interpretation. The p-value of each variable (Var.pval) is based on comparing the deviance explained by the unpermuted model (Model.dev.exp) to the distribution of deviance explained when the focal variable is permuted. Similarly, the model p-value (Model.pval) is based on a comparison between the deviance explained by the unpermuted model and the distribution of deviance explained when the entire predictor matrix is permuted.
